In-silico molecular enrichment and clearance of the human intracranial space

### 2025-02-25 paper/main.tex
Updates from Overleaf

\subsection*{In-silico predictions of intracranial molecular enrichment and clearance after intrathecal injection}
\begin{figure}
    \centering
    \includegraphics[width=0.95\textwidth]{figures/figure1.png}
     \caption{A) Illustration of the spatial domains of the model: CSF-filled space (ventricles and SAS) $\Omega_{\rm CSF}$, parenchyma $\Omega_{\rm PAR}$ and the PVS surrounding arteries $\Lambda_{\rm artery}$ and the veins $\Lambda_{\rm vein}$ and their interfaces and boundaries (see Methods); B) 3D rendering of the actual geometry used in this study parenchyma, CSF (clipped, blue), arterial network (red), venous network (blue)); C) arterial network (red) and venous network (blue); D) illustration of the modelled clinical tracer injection procedure; E) time of simulated first tracer arrival after intrathecal injection (concentration exceeds $0.1\,$mol/l); F) time of first tracer arrival in the arterial PVS (top) and venous PVS (bottom); G) sagittal, coronal and axial view of computed tracer concentrations over T1-image 1, 6, 12 and 24\,h after injection (low concentrations transparent, pial surface in pink, arachnoid membrane in cyan, and arteries in dark red); H) total amount of tracer in each cranial compartment over the first 24\,h; I) average tracer concentration in each compartment over the first 24\,h. }
     \label{fig:results1}
\end{figure}
Using previously published multi-modal magnetic resonance imaging
(MRI) data~\cite{hodneland2019new, deistung2009tof,
  schweser2012quantitative, reichenbach2012future,
  deistung2017overview}, we construct a multiscale computational
representation of the human intracranial compartments consisting of
the CSF spaces and brain parenchyma as three-dimensional (3D) domains
and with the PVSs surrounding major surface arteries and veins as
embedded networks of topologically one-dimensional (1D) curves
(\Cref{fig:results1}A--C). We consider a solute concentration field,
varying in space and time, in the 3D domains and in the PVS networks,
and assume that the solute can cross between these compartments
through semi-permeable membranes. As the drivers and modes of
intracranial transport are under substantial
debate\cite{smith2019going, proulx2021cerebrospinal,
  bohr2022glymphatic, hladky2022glymphatic, betsholtz2024advances},
our first target is to establish a baseline model accounting for a
reasonably conservative set of mechanisms and their integrated effect
over a timescale of several minutes to a few days. To this end, we
assume that the solute will (i) diffuse within all compartments, with
diffusivity depending on the effective properties of the relevant
medium\cite{sykova2008diffusion}; (ii) experience significant
dispersive effects due to the pulsatile flow of CSF induced by the
cardiac and respiratory cycles\cite{vinje2019respiratory,
  sharp2019dispersion, ray2021quantitative, troyetsky2021dispersion};
and (iii) be convected by a (small) net flow of CSF resulting from
production in the choroid plexus with CSF efflux across the upper
convexity\cite{hornkjol2022csf} and from the peristaltic pumping
effect of pulse wave pulsations in surface periarterial
spaces~\cite{mestre2018flow, gjerde2023directional}. Mathematically, this model is represented by a
mixed-dimensional system of coupled time-dependent partial
differential equations\cite{masri2024modelling}, which we solve
numerically with high accuracy using a mass-conserving finite element
scheme and the FEniCS finite element software\cite{alnaes2015fenics,
  kuchta2020assembly} (see Methods). The computational framework and
associated software are all openly available~\cite{ZENODO}.

@@

The PVSs are recognized across species as critical pathways for solute
transport in and around the brain\cite{}, and thus as potential
targets for enhancing brain drug delivery and metabolic waste
clearance\cite{}. However, whether CSF flows more rapidly in PVSs and
what the forces and mechanisms required to drive such flow are, remain
as key points of debate~\cite{bohr2022glymphatic,
  van2024caa}. Motivated by experimental observations in animal
models~\cite{iliff2012paravascular, iliff2013cerebral, mestre2018flow,
  bedussi2018paravascular}, there is now a remarkable body of
literature on modelling perivascular fluid flow and
transport~\cite{bilston2003arterial, asgari2016glymphatic,
  rey2018pulsatile, daversin2020mechanisms, sharp2019dispersion,
  thomas2019fluid, kedarasetti2020functional, kedarasetti2020arterial,
  troyetsky2021dispersion, martinac2021phase, gjerde2023directional,
  nozaleda2024arterial}. Here, we ask how these proposed mechanisms
would translate from idealized geometries to human vascular networks
and moreover evaluate their integrated effect in the context of
intracranial solute transport.

Our periarterial network extends from the internal carotid arteries
and basilar artery through up to 18 bifurcations to reach upstream
network ends located within the SAS or up to 6 mm inside the
parenchyma (\Cref{fig:results1}A--C, \Cref{fig:pvs_a}A) and includes
major surface arteries of radius 0.5--1.4 mm. Imposing the pressure
field induced by CSF production at the periarterial network ends
induces slow steady CSF flow of variable direction in these PVSs, with
an average velocity of 0.08 \textmu m/s (antegrade) and a maximum
velocity of 0.55 \textmu m/s (\Cref{fig:pvs_a}B). On the other hand,
traveling waves of arterial wall motion (\Cref{fig:pvs_a}C), such as
the pulse wave or other vasomotion~\cite{vanveluw2020vasomotion,
  munting2023spontaneous, bojarskaite2023sleep,
  broggini2024long,hauglund2025norepinephrine}, also induce net
directional flow in the PVS~\cite{kedarasetti2020arterial,
  kedarasetti2020functional, coenen2021lubrication,
  gjerde2023directional, nozaleda2024arterial} -- of magnitude and
direction depending on the amplitude, frequency, and length of the
waves and the characteristics of the perivascular
network~\cite{gjerde2023directional}. Applying a semi-analytic model
of the net flow induced by peristalsis in perivascular
networks~\cite{gjerde2023directional} (see Methods), we estimate that
the cardiac pulse wave alone, traveling at a frequency of 1 Hz with a
wavelength of 2.0 m and a 1\% wall displacement~\cite{jung2021novel},
will induce mainly antegrade PVS flow with an average net velocity of
0.92 \textmu m/s while reaching up to 7.31 \textmu m/s near larger,
ventral vessels such as the MCA (\Cref{fig:pvs_a}D). The same theory
predicts that strong ultraslow vasomotions, if traveling antegrade at
0.1Hz with a wavelength of 0.02 m and a 10\%
displacement~\cite{broggini2024long}, will induce both retrograde and
antegrade net PVS flow with an average velocity of 13.05 \textmu m/s
and maximum velocity 54.44 \textmu m/s (\Cref{fig:pvs_a}E).

@@

\begin{figure}
    \centering
    \includegraphics[width=0.95\textwidth]{figures/figure3a.png}
    \caption{\textbf{A static pressure gradient and peristaltic pumping are potential drivers of perivascular flow.}
    A) Surface arterial network colored by arterial radius with vessel locations labeled;
    B) Net PVS flow induced by CSF production (cf.~\Cref{fig:csf}D) -- histograms show the relative frequencies of local periarterial velocities. Positive values indicate antegrade PVS flow, while negative values indicate retrograde PVS flow;
    C) Illustration of the concept of peristaltic pumping;
    D) Estimated net PVS flow induced by pulse wave peristaltic pumping;
    E) Estimated net PVS flow induced by vasomotion/slow wave peristaltic pumping.}
    \label{fig:pvs_a}
\end{figure}

>>>>

\subsection*{In-silico predictions of intracranial molecular enrichment and clearance after intrathecal injection}
\begin{figure}
    \centering
    \includegraphics[width=0.93\textwidth]{figures/figure1.png}
     \caption{
     \textbf{Our in-silico model predicts tracer transport in PVS, SAS and parenchyma following intrathecal injection.}\\
     A) Illustration of the spatial domains of the model: CSF-filled space (ventricles and SAS) $\Omega_{\rm CSF}$, parenchyma $\Omega_{\rm PAR}$ and the PVS surrounding arteries $\Lambda_{\rm artery}$ and the veins $\Lambda_{\rm vein}$ and their interfaces and boundaries (see Methods); B) 3D rendering of the actual geometry used in this study parenchyma, CSF (clipped, blue), arterial network (red), venous network (blue)); C) arterial network (red) and venous network (blue); D) illustration of the modelled clinical tracer injection procedure; E) time of simulated first tracer arrival after intrathecal injection (concentration exceeds $0.1\,$mol/l); F) time of first tracer arrival in the arterial PVS (top) and venous PVS (bottom); G) sagittal, coronal and axial view of computed tracer concentrations over T1-image 1, 6, 12 and 24\,h after injection (low concentrations transparent, pial surface in pink, arachnoid membrane in cyan, and arteries in dark red); H) total amount of tracer in each cranial compartment over the first 24\,h; I) average tracer concentration in each compartment over the first 24\,h. }
     \label{fig:results1}
\end{figure}
Using previously published multi-modal magnetic resonance imaging
(MRI) data~\cite{hodneland2019new, deistung2009tof,
  schweser2012quantitative, reichenbach2012future,
  deistung2017overview}, we construct a multiscale computational
representation of the human intracranial compartments consisting of
the CSF spaces and brain parenchyma as three-dimensional (3D) domains
and with the PVSs surrounding major surface arteries and veins as
embedded networks of topologically one-dimensional (1D) curves
(\Cref{fig:results1}A--C). We consider a solute concentration field,
varying in space and time, in the 3D domains and in the PVS networks,
and assume that the solute can cross between these compartments
through semi-permeable membranes. As the drivers and modes of
intracranial transport are under substantial
debate\cite{smith2019going, proulx2021cerebrospinal,
  bohr2022glymphatic, hladky2022glymphatic, betsholtz2024advances},
our first target is to establish a baseline model accounting for a
reasonably conservative set of mechanisms and their integrated effect
over a timescale of several minutes to a few days. To this end, we
assume that the solute will (i) diffuse within all compartments, with
diffusivity depending on the effective properties of the relevant
medium\cite{sykova2008diffusion}; (ii) experience significant
dispersive effects due to the pulsatile flow of CSF induced by the
cardiac and respiratory cycles\cite{vinje2019respiratory,
  sharp2019dispersion, ray2021quantitative, troyetsky2021dispersion};
and (iii) be convected by a (small) net flow of CSF resulting from
production in the choroid plexus with CSF efflux across the upper
convexity\cite{hornkjol2022csf} and from the peristaltic pumping
effect of pulse wave pulsations in surface periarterial
spaces~\cite{mestre2018flow, gjerde2023directional}. Mathematically, this model is represented by a
mixed-dimensional system of coupled time-dependent partial
differential equations\cite{masri2024modelling}, which we solve
numerically with high accuracy using a mass-conserving finite element
scheme and the FEniCS finite element software\cite{alnaes2015fenics,
  kuchta2020assembly} (see Methods). The computational framework and
associated software are all openly available~\cite{ZENODO}.

@@

The PVSs are recognized across species as critical pathways for solute
transport in and around the brain\cite{}, and thus as potential
targets for enhancing brain drug delivery and metabolic waste
clearance\cite{}. However, whether CSF flows more rapidly in PVSs and
what the forces and mechanisms required to drive such flow are, remain
as key points of debate~\cite{bohr2022glymphatic,
  van2024caa}. Motivated by experimental observations in animal
models~\cite{iliff2012paravascular, iliff2013cerebral, mestre2018flow,
  bedussi2018paravascular}, there is now a remarkable body of
literature on modelling perivascular fluid flow and
transport~\cite{bilston2003arterial, asgari2016glymphatic,
  rey2018pulsatile, daversin2020mechanisms, sharp2019dispersion,
  thomas2019fluid, kedarasetti2020functional, kedarasetti2020arterial,
  troyetsky2021dispersion, martinac2021phase, gjerde2023directional,
  nozaleda2024arterial}. Here, we ask how these proposed mechanisms
would translate from idealized geometries to human vascular networks
and moreover, evaluate their integrated effect in the context of
intracranial solute transport.

Our periarterial network extends from the internal carotid arteries
and basilar artery through up to 18 bifurcations to reach upstream
network ends located within the SAS or up to 6 mm inside the
parenchyma (\Cref{fig:results1}A--C, \Cref{fig:pvs_a}A) and includes
major surface arteries of radius 0.5--1.4 mm. Imposing the pressure
field induced by CSF production at the periarterial network ends
induces slow steady CSF flow of variable direction in these PVSs, with
an average velocity of 0.08 \textmu m/s (antegrade) and a maximum
velocity of 0.55 \textmu m/s (\Cref{fig:pvs_a}B). On the other hand,
traveling waves of arterial wall motion (\Cref{fig:pvs_a}C), such as
the pulse wave or other vasomotion~\cite{vanveluw2020vasomotion,
  munting2023spontaneous, bojarskaite2023sleep,
  broggini2024long,hauglund2025norepinephrine}, also induce net
directional flow in the PVS~\cite{kedarasetti2020arterial,
  kedarasetti2020functional, coenen2021lubrication,
  gjerde2023directional, nozaleda2024arterial} -- of magnitude and
direction depending on the amplitude, frequency, and length of the
waves and the characteristics of the perivascular
network~\cite{gjerde2023directional}. Applying a semi-analytic model
of the net flow induced by peristalsis in perivascular
networks~\cite{gjerde2023directional} (see Methods), we estimate that
the cardiac pulse wave alone, travelling at a frequency of 1 Hz with a
wavelength of 2.0 m and a 1\% wall displacement~\cite{jung2021novel},
will induce mainly antegrade PVS flow with an average net velocity of
0.92 \textmu m/s while reaching up to 7.31 \textmu m/s near larger,
ventral vessels such as the MCA (\Cref{fig:pvs_a}D,E). The same theory
predicts that strong ultraslow vasomotions, if travelling antegrade at
0.1Hz with a wavelength of 0.02 m and a 10\%
displacement~\cite{broggini2024long}, will induce both retrograde and
antegrade net PVS flow with an average velocity of 13.05 \textmu m/s
and maximum velocity 54.44 \textmu m/s (\Cref{fig:pvs_a}D,F).

@@

\begin{figure}
    \centering
    \includegraphics[width=0.95\textwidth]{figures/figure3a.png}
    \caption{\textbf{A static pressure gradient and peristaltic pumping are potential drivers of perivascular flow.}
    A) Surface arterial network coloured by arterial radius with vessel locations labelled;
    B) Net PVS flow induced by CSF production (cf.~\Cref{fig:csf}
    -- histograms show the relative frequencies of local periarterial velocities. Positive values indicate antegrade PVS flow, while negative values indicate retrograde PVS flow;
    C) Illustration of the concept of peristaltic pumping;
    D) histograms of estimated net PVS flow induced by pulse wave or vasomotion peristaltic pumping;
    E) Estimated net PVS flow induced by pulse wave peristaltic pumping.
    F) Estimated net PVS flow induced by vasomotion/slow wave peristaltic pumping.}
    \label{fig:pvs_a}
\end{figure}
